---
title: "Calculating absolute abundances for frog and fly"
author: "Edward Cruz"
date: "2025-09-24"
output: github_document
---


```{r Libraries }

rm(list=ls(all=T))

library(Biostrings)
library(stringi)
library(arrow)
library(tidyr)
library(dplyr)
library(stringr)
library(Peptides)
library(ggplot2)

knitr::opts_chunk$set(fig.path = "figures/turnover_abs_quant/")

```


```{r Importing fasta files and calculating molecular weights }

#FASTA file used to search data
frog_fasta <- readAAStringSet("Files/FASTA/XENLA_longest-transcript_10p1_GFY-Fix.fasta")

#Converting to a dataframe for merging
frog_AA.df <- data.frame(
  Protein_ID = names(frog_fasta),
  Sequence = as.character(frog_fasta),
  stringsAsFactors = FALSE )
row.names(frog_AA.df) <- NULL
frog_AA.df["Protein_ID"] <- sapply(strsplit(frog_AA.df$`Protein_ID`, split = " "),
                                   "[", 1)

#FASTA file used to search data
fly_fasta <- readAAStringSet("Files/Reference/Dmel_Uniprot_Proteome_092722.fasta")

#Converting to a dataframe for merging
fly_AA.df <- data.frame(
  Protein_ID = names(fly_fasta),
  Sequence = as.character(fly_fasta),
  stringsAsFactors = FALSE )
row.names(fly_AA.df) <- NULL
fly_AA.df["Protein_ID"] <- sapply(strsplit(fly_AA.df$`Protein_ID`, split = " "),
                                  "[", 1)
fly_AA.df["Protein_ID"] <- sapply(strsplit(fly_AA.df$Protein_ID, split = "\\|"),
                                  "[", 2)
#---- Calculating MW ----------

frog_AA.df["Sequence"] <- str_remove_all(frog_AA.df$Sequence,
                                         "(?i)[^ARNDCEQGHILKMFPSTWYV]")
fly_AA.df["Sequence"] <- str_remove_all(fly_AA.df$Sequence,
                                        "(?i)[^ARNDCEQGHILKMFPSTWYV]")

frog_AA.df["kDa"] <- sapply(frog_AA.df$Sequence, function(seq){
    return(mw(seq)/1000) })
fly_AA.df["kDa"] <- sapply(fly_AA.df$Sequence, function(seq){
    return(mw(seq)/1000) })

frog_AA.df <- frog_AA.df %>% select(Protein_ID, kDa, Sequence)
fly_AA.df <- fly_AA.df %>% select(Protein_ID, kDa, Sequence)

```


```{r Function for in silico digest }

tryptic_digest <- function(fasta_file) {

  #FASTA file used to search data
  protein_fasta <- 
    readAAStringSet(fasta_file)
  
  #Converting to a dataframe for merging
  protein_seq.df <- data.frame(
    Protein_ID = names(protein_fasta),
    Sequence = as.character(protein_fasta),
    stringsAsFactors = FALSE )
  row.names(protein_seq.df) <- NULL
  
  #Setting names and sequences for contaminants
  contams_seq.df <- protein_seq.df[grepl("contaminant",
                                         protein_seq.df$Protein_ID),]
  contams_seq.df["Protein_ID"] <- sapply(strsplit(contams_seq.df$Protein_ID, split = "\\|"),
                                         "[", 2)
  contams_seq.df["Description"] <- rep("contaminant", nrow(contams_seq.df))
  
  #Setting names for fwd sequences
  protein_seq.df <- protein_seq.df[!grepl("contaminant",
                                          protein_seq.df$Protein_ID),]
  protein_seq.df["Description"] <- rep("Temp", nrow(protein_seq.df))
  
  #Merging contaminants and organizing data.frame
  protein_seq.df <- rbind(contams_seq.df, protein_seq.df)
  
  #Determining number of tryptic peptides in sample
  protein_seq.df <- cbind(protein_seq.df,
                          lapply(protein_seq.df$Sequence, function(seq){
    
    peps <- stri_split_regex(seq, "(?<=[RK])")[[1]]  
    peps <- peps[nchar(peps) >= 7 & nchar(peps) <= 30]
    
    theo_peps <- length(peps)
    peps <- paste(peps, collapse = ",")
    
    outline <- data.frame(Seq_Length = nchar(seq),
                          Theo_Peps = theo_peps,
                          Fragments = peps)
    
    return(outline) }) %>% bind_rows )
  
  protein_seq.df <- protein_seq.df %>% 
    select(Protein_ID, Description, Seq_Length, Theo_Peps,
           Sequence, Fragments)
  
  return(protein_seq.df) }

```


```{r In silico digestion of frog proteome }

XLA_Digest <- 
  tryptic_digest("Files/FASTA/DIA-NN/XENLA_longest-transcript_10p1_GFY-Fix_w_contams.fasta")

XLA_Digest[!XLA_Digest$Description=="contaminant", "Description"] <-
  sapply(strsplit(XLA_Digest[!XLA_Digest$Description=="contaminant",]$`Protein_ID`,
                  split = " "), "[", 2)
XLA_Digest[is.na(XLA_Digest$Description), "Description"] <- "N/A"

XLA_Digest[!XLA_Digest$Description=="contaminant", "Protein_ID"] <-
  sapply(strsplit(XLA_Digest[!XLA_Digest$Description=="contaminant",]$`Protein_ID`,
                  split = " "), "[", 1)

```


```{r In silico digestion of fly proteome }

Dmel_Digest <- 
  tryptic_digest("Files/FASTA/DIA-NN/Dmel_Uniprot_Proteome_092722_w_contams.fasta")

Dmel_Digest[!Dmel_Digest$Description=="contaminant", "Description"] <-
  sapply(Dmel_Digest[!Dmel_Digest$Description=="contaminant",]$Protein_ID,
         function(x){
           description <- str_match(x, " ([^\r\n]*?)(?=\\s*OS=)")[,2]
         return(description) })
Dmel_Digest[!Dmel_Digest$Description=="contaminant", "Protein_ID"] <- 
  sapply(strsplit(Dmel_Digest[!Dmel_Digest$Description=="contaminant",]$Protein_ID,
                  split = "\\|"), "[", 2)

head(Dmel_Digest)

```


```{r Importing DIANN data }

read_DIANN <- function(file_name) {

  #Selecting proteins that meet q-value threshold
  df <- read_parquet(file_name)
  df <- df[df$Lib.PG.Q.Value <= 0.01,]

  # For a Protein.Group to be listed by itself, it therefore requires at
  #     least one peptide that can UNIQUELY map to it.  
  df <- df[!grepl("contam", df$Protein.Names),]
  df <- df[!is.na(df$Precursor.Quantity),]
  df <- df[!df$Precursor.Quantity==0,]

  df <- df[c("Run", "Protein.Group", "Genes", "Stripped.Sequence",
             "Precursor.Quantity")]
  df <- df %>% 
    group_by(Run, `Protein.Group`, Genes, `Stripped.Sequence`) %>%
    dplyr::summarise(`Precursor.Quantity` = sum(`Precursor.Quantity`),
                     .groups = "drop") %>%
    as.data.frame

  return(df) }


XLA_Abs_DIANN <- read_DIANN("Data/DIA/Frog_AbsQ_report.parquet")
Dmel_Abs_DIANN <- read_DIANN("Data/DIA/Fly_AbsQ_report.parquet")

head(XLA_Abs_DIANN)
head(Dmel_Abs_DIANN)

```


```{r Merging and filtering DIA results  }

merge_DIA_FASTA <- function(data_df, fasta_df){
  
  all_data <- data_df %>%
    #Exploding dataframe by treating all ";" as individual entries
    mutate(ProteinToken = strsplit(`Protein.Group`, ";", fixed = TRUE)) %>%
    tidyr::unnest(ProteinToken) %>% #Unnesting exploded df
    #Merging individual tokens with FASTA file
    inner_join(fasta_df[c("Protein_ID", "Theo_Peps", "Seq_Length")],
               by = c("ProteinToken" = "Protein_ID")) %>% 
    
    #Grouping by what we started with and determining med,min,max
    #     for all peptides in Protein.Group
    group_by(across(all_of(names(data_df)))) %>%
    dplyr::summarise(
      Theo_Peps_Median = median(Theo_Peps, na.rm = TRUE),
      Theo_Peps_Min = min(Theo_Peps, na.rm = TRUE),
      Theo_Peps_Max = max(Theo_Peps, na.rm = TRUE),
      Length_Median = median(Seq_Length, na.rm = TRUE),
      Length_Min = min(Seq_Length, na.rm = TRUE),
      Length_Max = max(Seq_Length, na.rm = TRUE),
      .groups = "drop") %>%
    
    #Getting protein groups that are within 20% spread in theoretical peptides
    filter(
      Theo_Peps_Max <= 1.2 * Theo_Peps_Min,
      Length_Max <= 1.2 * Length_Min) %>%

    #Returning back original columns and modified theo_peps and sequence length    
    transmute(
      !!!syms(names(data_df)),
      Theo_Peps = Theo_Peps_Median,
      Length = Length_Median)    
  
  return(all_data) }

XLA_Abs_FASTA_merge <- merge_DIA_FASTA(XLA_Abs_DIANN, XLA_Digest)
Dmel_Abs_FASTA_merge <- merge_DIA_FASTA(Dmel_Abs_DIANN, Dmel_Digest)

head(XLA_Abs_FASTA_merge)
head(Dmel_Abs_FASTA_merge)

```


```{r Yolk proteins set apart in individual compartment }

fly_yolk_set <- c("P02843", "P02844", "P06607") #Yp1, Yp2, Yp3
names(fly_yolk_set) <- c("Yp1", "Yp2", "Yp3")

frog_yolk_set <- c("XBmRNA35014|XBXL10_1g18990", "XBmRNA39801|XBXL10_1g21498",
                   "XBmRNA35013|XBXL10_1g18989", "XBmRNA35012|XBXL10_1g18988",
                   "XBmRNA67639|XBXL10_1g35803")
names(frog_yolk_set) <- c("vtgb1.L", "vtgb1.S", "vtga1.L", "vtga2.L", "serpina")

```


```{r Functions for calculating iBAQ values and unnesting data }

Abs_Conc_per_run <- function(data_df, run_no, cell_conc) {
  sub_df <- data_df[data_df$Run == run_no,]
  sub_df <- sub_df %>% group_by(`Protein.Group`, Genes, Theo_Peps) %>%
    dplyr::summarise(Sum_Pep_Int = sum(`Precursor.Quantity`),
                     .groups = "drop")

  sub_df <- sub_df[!sub_df$Theo_Peps == 0,] #Removing weird proteins
  
  #Estimating molar units based on peptides identified    
  sub_df["iBAQ_Value"] <- sub_df$Sum_Pep_Int / sub_df$Theo_Peps

  return(sub_df) }

unnest_data <- function(data_df) {
  
  data_df <- data_df %>%
      #Exploding dataframe by treating all ";" as individual entries
      mutate(Ind_Protein_ID = strsplit(`Protein.Group`, ";", fixed = TRUE)) %>%
      unnest(Ind_Protein_ID) %>% #Unnesting exploded df
      select(Ind_Protein_ID, everything())
  
  return(data_df) }

```


```{r Applying functions to all runs }

XLA_T5_RT0_Abs <- Abs_Conc_per_run(XLA_Abs_FASTA_merge, 'ORC_05940_Frog_T5-RT0', 2)
XLA_T5_N12_Abs <- Abs_Conc_per_run(XLA_Abs_FASTA_merge, 'ORC_05942_Frog_T5-N12', 2)
XLA_T6_RT0_Abs <- Abs_Conc_per_run(XLA_Abs_FASTA_merge, 'ORC_05944_Frog_T6-RT0', 2)
XLA_T6_N12_Abs <- Abs_Conc_per_run(XLA_Abs_FASTA_merge, 'ORC_05946_Frog_T6-N12', 2)

XLA_T5_RT0_Abs <- unnest_data(XLA_T5_RT0_Abs)
XLA_T5_N12_Abs <- unnest_data(XLA_T5_N12_Abs)
XLA_T6_RT0_Abs <- unnest_data(XLA_T6_RT0_Abs)
XLA_T6_N12_Abs <- unnest_data(XLA_T6_N12_Abs)

head(XLA_T5_RT0_Abs)

#-------------------

Dmel_T6_T0A_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05948_Fly_T6-T0A', 1.6)
Dmel_T6_T0B_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05950_Fly_T6-T0B', 1.6)
Dmel_T6_N16_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05952_Fly_T6-N16', 1.6)
Dmel_T7_T0A_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05954_Fly_T7-T0A', 1.6)
Dmel_T7_T0B_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05956_Fly_T7-T0B', 1.6)
Dmel_T7_N16_Abs <- Abs_Conc_per_run(Dmel_Abs_FASTA_merge, 'ORC_05958_Fly_T7-N16', 1.6)

Dmel_T6_T0A_Abs <- unnest_data(Dmel_T6_T0A_Abs)
Dmel_T6_T0B_Abs <- unnest_data(Dmel_T6_T0B_Abs)
Dmel_T6_N16_Abs <- unnest_data(Dmel_T6_N16_Abs)
Dmel_T7_T0A_Abs <- unnest_data(Dmel_T7_T0A_Abs)
Dmel_T7_T0B_Abs <- unnest_data(Dmel_T7_T0B_Abs)
Dmel_T7_N16_Abs <- unnest_data(Dmel_T7_N16_Abs)

head(Dmel_T6_T0A_Abs)

```


Citation for frog absolute proteomics:

Gurdon and Wickens // Evi's Paper

280 ug yolk protein 
45 ug non-yolk protein

```{r Splitting yolk and nonyolk frog protein for absolute estimation }

split_frog_yolk <- function(df, yolk_amount, nonyolk_amount) {

  df <- merge(df, frog_AA.df, by.x="Ind_Protein_ID", by.y="Protein_ID")

  sub_yolk <- df[df$Ind_Protein_ID %in% frog_yolk_set,]
  sub_nonyolk <- df[!df$Ind_Protein_ID %in% frog_yolk_set,]
  
  #Calculate amount of protein in milligrams
  sub_yolk["iBAQ_WeightFraction"] <- sub_yolk$iBAQ_Value * sub_yolk$kDa #mult by kda for weight
  #Calculate mass fraction
  sub_yolk["iBAQ_WeightFraction"] <- sub_yolk$iBAQ_WeightFraction / sum(sub_yolk$iBAQ_WeightFraction)  
  sub_yolk["iBAQ_Weight_mg"] <- sub_yolk$iBAQ_WeightFraction * (yolk_amount/1000) #mg calc
  #Calculate mg/mL then divide by kDa -> mM
  sub_yolk["iBAQ_Conc"] <- sub_yolk$iBAQ_Weight_mg / 0.001  
  sub_yolk["iBAQ_Conc"] <- sub_yolk$iBAQ_Conc / sub_yolk$kDa  
      
  #--------------

  #Calculate amount of protein in milligrams
  sub_nonyolk["iBAQ_WeightFraction"] <- sub_nonyolk$iBAQ_Value * sub_nonyolk$kDa #mult by kda for weight
  #Calculate mass fraction
  sub_nonyolk["iBAQ_WeightFraction"] <- sub_nonyolk$iBAQ_WeightFraction /
    sum(sub_nonyolk$iBAQ_WeightFraction)  
  sub_nonyolk["iBAQ_Weight_mg"] <- sub_nonyolk$iBAQ_WeightFraction * (nonyolk_amount/1000) #mg calc
  #Calculate mg/mL then divide by kDa -> mM
  sub_nonyolk["iBAQ_Conc"] <- sub_nonyolk$iBAQ_Weight_mg / 0.001  
  sub_nonyolk["iBAQ_Conc"] <- sub_nonyolk$iBAQ_Conc / sub_nonyolk$kDa  

  #--------------
  
  recalc_df <- rbind(sub_yolk, sub_nonyolk)
  recalc_df <- recalc_df %>% arrange(desc(iBAQ_Conc))
  recalc_df[c("iBAQ_WeightFraction", "iBAQ_Weight_mg")] <- NULL

  recalc_df <- recalc_df %>% 
    relocate(Sequence, .after = last_col())  
    
  return(recalc_df) }

#These are from Evi's paper assuming:
#  ~ non-yolk protein amount doesn't increase until after NF30
XLA_T5_RT0_Abs <- split_frog_yolk(XLA_T5_RT0_Abs, 280, 45)
XLA_T5_N12_Abs <- split_frog_yolk(XLA_T5_N12_Abs, 280, 50)
XLA_T6_RT0_Abs <- split_frog_yolk(XLA_T6_RT0_Abs, 280, 45)
XLA_T6_N12_Abs <- split_frog_yolk(XLA_T6_N12_Abs, 280, 50)

head(XLA_T5_RT0_Abs)

```


Citation for fly absolute concentrations:

110 mg of protein in mature oocyte
80 mg at Stage 5
60 mg at hatching

https://ijdb.ehu.eus/article/pdf/2518452
https://link.springer.com/content/pdf/10.1007/BF00848689.pdf

-> Total protein decreases by ~25% at hatching from window start

1.8 ug total protein at start
1.35 ug total protein at end

Trying:
0.8 ug total protein at start
0.6 ug total protein at end


```{r Splitting yolk and nonyolk fly protein for absolute estimation }

fly_yolk_set <- c("P02843", "P02844", "P06607") #Yp1, Yp2, Yp3
names(fly_yolk_set) <- c("Yp1", "Yp2", "Yp3")

split_fly_yolk <- function(df, protein_amount) {

  df <- merge(df, fly_AA.df, by.x="Ind_Protein_ID", by.y="Protein_ID")

  #Calculate amount of protein in milligrams
  df["iBAQ_WeightFraction"] <- df$iBAQ_Value * df$kDa #mult by kda for weight
  #Calculate mass fraction
  df["iBAQ_WeightFraction"] <- df$iBAQ_WeightFraction / sum(df$iBAQ_WeightFraction)
  df["iBAQ_Weight_mg"] <- df$iBAQ_WeightFraction * (protein_amount/1000) #mg calc
  #Calculate mg/mL then divide by kDa -> mM
  df["iBAQ_Conc"] <- df$iBAQ_Weight_mg / 0.00001
  df["iBAQ_Conc"] <- df$iBAQ_Conc / df$kDa
  df <- df %>% arrange(desc(iBAQ_Conc))

  df[c("iBAQ_WeightFraction", "iBAQ_Weight_mg")] <- NULL
  df <- df %>% relocate(Sequence, .after = last_col())

  return(df) }

Dmel_T6_T0A_Abs <- split_fly_yolk(Dmel_T6_T0A_Abs, 0.8)
Dmel_T6_T0B_Abs <- split_fly_yolk(Dmel_T6_T0B_Abs, 0.8)
Dmel_T6_N16_Abs <- split_fly_yolk(Dmel_T6_N16_Abs, 0.6) #total decrease by 25%
Dmel_T7_T0A_Abs <- split_fly_yolk(Dmel_T7_T0A_Abs, 0.8)
Dmel_T7_T0B_Abs <- split_fly_yolk(Dmel_T7_T0B_Abs, 0.8)
Dmel_T7_N16_Abs <- split_fly_yolk(Dmel_T7_N16_Abs, 0.6) #total decrease by 25%

```


```{r Combining runs for representative dataset }

merge_datasets <- function(data1, data2, l1, l2) {

  protein_counts <- c(data1$Ind_Protein_ID, data2$Ind_Protein_ID)  
  protein_counts <- data.frame(table(protein_counts))
  
  single_ids <- protein_counts[protein_counts$Freq==1,]
  multiple_ids <- protein_counts[protein_counts$Freq>1,]
    
  no_reps <- rbind(data1[data1$Ind_Protein_ID %in% single_ids[[1]],],
                   data2[data2$Ind_Protein_ID %in% single_ids[[1]],])
  no_reps <- no_reps[c("Ind_Protein_ID", "iBAQ_Conc", "Sequence")]
  no_reps["Type"] <- "Point"
  
  reps <- merge(data1[data1$Ind_Protein_ID %in% multiple_ids[[1]],
                      c("Ind_Protein_ID", "iBAQ_Conc")],
                data2[data2$Ind_Protein_ID %in% multiple_ids[[1]],
                      c("Ind_Protein_ID", "iBAQ_Conc", "Sequence")],
                by="Ind_Protein_ID", suffixes=c(paste("_", c(l1, l2), sep="")),
                all=TRUE)

  reps["iBAQ_Conc"] <- (reps[[2]] + reps[[3]])/2
  reps["Type"] <- "Interval"

  out_df <- rbind(reps[c("Ind_Protein_ID", "iBAQ_Conc",
                         "Type", "Sequence")],
                  no_reps[c("Ind_Protein_ID", "iBAQ_Conc",
                            "Type", "Sequence")])

  return(out_df) }


XLA_RT0_Abs <- merge_datasets(XLA_T5_RT0_Abs, XLA_T6_RT0_Abs, "T5", "T6")
XLA_N12_Abs <- merge_datasets(XLA_T5_N12_Abs, XLA_T6_N12_Abs, "T5", "T6")

Dmel_T0_Abs <-
  merge_datasets(merge_datasets(Dmel_T6_T0A_Abs, Dmel_T6_T0B_Abs, "A", "B"),
                 merge_datasets(Dmel_T7_T0A_Abs, Dmel_T7_T0B_Abs, "A", "B"),
                 "T6", "T7")
Dmel_N16_Abs <- merge_datasets(Dmel_T6_N16_Abs, Dmel_T7_N16_Abs, "T6", "T7")

```


```{r Exporting data for use in external scripts }

write.csv(XLA_RT0_Abs, "Files/AbsC/XLA_RT0_Abs-Concentrations.csv",
          row.names = FALSE)
write.csv(XLA_N12_Abs, "Files/AbsC/XLA_N12_Abs-Concentrations.csv",
          row.names = FALSE)

write.csv(Dmel_T0_Abs, "Files/AbsC/Dmel_T0_Abs-Concentrations.csv",
          row.names = FALSE)
write.csv(Dmel_N16_Abs, "Files/AbsC/Dmel_N16_Abs-Concentrations.csv",
          row.names = FALSE)

```


```{r Importing frog annotation }

human_map <- read.csv("Files/Reference/Xen10_to_Human_Name_Assign.csv")
human_map["Protein_ID"] <- sapply(human_map$Protein_ID, function(x){
  split_id <- strsplit(x, "\\|")[[1]]
  reformat <- paste0(split_id[3], "|", split_id[2])

  return(reformat) })

human_map <- human_map[c("Protein_ID", "XLA_Gene", "Human_Gene", "Description")]

```


```{r Importing fly annotation }

Dmel.genes <- read.csv("Files/Reference/Dmel_GeneTable_Descriptions.csv")
colnames(Dmel.genes)[1] <- "Protein_ID"
Dmel.genes <- Dmel.genes[c("Protein_ID", "Gene_Symbol", "Name", "Description")]

```


```{r Exporting supplemental frog table }

supp_df <- XLA_RT0_Abs
supp_df["iBAQ_Conc_uM"] <- supp_df$iBAQ_Conc * 1000

supp_df <- merge(human_map, supp_df[c("Ind_Protein_ID", "iBAQ_Conc_uM")],
                 by.x="Protein_ID", by.y="Ind_Protein_ID") %>%
  arrange(desc(iBAQ_Conc_uM))

dir.create("Files/Supp_Tables", showWarnings = FALSE, recursive = TRUE)
write.csv(supp_df, "Files/Supp_Tables/Frog_Start-AbsQ_Supplementary_Table.csv",
          row.names = FALSE)

```


```{r Exporting supplemental fly table }

supp_df <- Dmel_T0_Abs
supp_df["iBAQ_Conc_uM"] <- supp_df$iBAQ_Conc * 1000
supp_df <- merge(Dmel.genes, supp_df[c("Ind_Protein_ID", "iBAQ_Conc_uM")],
                 by.x="Protein_ID", by.y="Ind_Protein_ID") %>%
  arrange(desc(iBAQ_Conc_uM))

dir.create("Files/Supp_Tables", showWarnings = FALSE, recursive = TRUE)
write.csv(supp_df, "Files/Supp_Tables/Fly_Start-AbsQ_Supplementary_Table.csv",
          row.names = FALSE)

```


```{r}

sessionInfo()

```


